Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis

### 2018-03-18 rigid_bodies.tex
Preparation for a short version

\section{Proposed Method}
\label{sec:methodology}

@@

The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion are \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
%
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
%
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
%
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
%
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$. In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion. Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity. Matrices $\mt M$ and $\mt I$ are diagonal ones. The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia. Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

@@

Description of analytical and NO\_SQUISH rotations...

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The equations of motion, in this case, are
\begin{subequations}
	\label{eq:ODE system for NVT}
	\begin{align}
%
\label{eq:nhc_r}
	&\dot{\vt r} =
	{\mt M}^{-1} {\vt p}, \\
%
\label{eq:nhc_p} 
	&\dot{\vt p} =
	{\vt F} - \alpha_1 {\vt p},\\
%
\label{eq:nhc_q}
	&\dot{\vt q} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
%
\label{eq:nhc_pi}
	&\dot{\vt \pi} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
%
\label{eq:nhc_eta}
	&\dot{\eta}_j = \alpha_j, \text{ and} \\
%
\label{eq:nhc_p_eta}
	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
	\end{align}
\end{subequations}

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where $Q_j$ is an inertial parameter for the corresponding thermostat. Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
\end{align}
\end{subequations}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set to zero at the onset or $L = 6N$ otherwise.\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, we make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain. The ``extended energy'' conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain. Of course, the NVE part can be integrated exactly as described before. The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\modified K$ to be scaled by the square of it, hence ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$. In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\modified K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$. It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.

@@

@@

For reasons that will become clear shortly, we employ an asymmetric, four-point stencil formula for estimating $\dot{\vt r}$ at a given instant $t$, which is
\begin{equation*}
\dot{\vt r}^{[3]} = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
\end{equation*}

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]} = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}. In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.

@@

@@

Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space. Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a point in phase space, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation}
\label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$). A generalized Liouville operator can then be defined as\cite{Sergi_2004}
\begin{equation*}
{\mathcal L} = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
\end{equation*}

>>>>

\section{Proposed Methodology}
\label{sec:methodology}

\subsection{NVE Dynamics: Notation and Symplectic Integration}

@@

The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion are \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
%
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
%
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
%
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
%
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$. In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion. Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity. Matrices $\mt M$ and $\mt I$ are diagonal ones. The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia. Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

@@

Brief description of analytical and NO-SQUISH rotations...

\subsection{NVT Dynamics with a Single Thermostat Chain}

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
\end{equation}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter. As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.

Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} Invaluable tools have been devised for the development and analysis of extended phase space methods. By employing the method of Sergi and Ferrario \cite{Sergi_2001}, we obtain the equations of motion for the NVT dynamics with a single thermostat chain, which are
\begin{subequations}
	\label{eq:ODE system for NVT}
	\begin{align}
%
\label{eq:nhc_r}
	&\dot{\vt r} =
	{\mt M}^{-1} {\vt p}, \\
%
\label{eq:nhc_p} 
	&\dot{\vt p} =
	{\vt F} - \alpha_1 {\vt p},\\
%
\label{eq:nhc_q}
	&\dot{\vt q} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \\
%
\label{eq:nhc_pi}
	&\dot{\vt \pi} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
%
\label{eq:nhc_eta}
	&\dot{\eta}_j = \alpha_j, \text{ and} \\
%
\label{eq:nhc_p_eta}
	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M.
	\end{align}
\end{subequations}

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$. Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
\end{align}
\end{subequations}

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$. It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$. Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates. Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$. In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$. In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.

\subsection{Measure-Preserving Integration for the NVT Dynamics}

As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. Ezra \cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion. In short, a splitting scheme will preserve a given measure if each individual propagator does it as well. By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain. Of course, the NVE part can be integrated exactly as described before. The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$. Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$.

@@

\subsection{Backward Error Analysis}

@@

For reasons that will become clear shortly, we employ an asymmetric, four-point stencil formula for estimating $\dot{\vt r}$ at a given instant $t$, which is
\begin{equation*}
\dot{\vt r}^{[3]}_t = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
\end{equation*}

In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
\begin{equation*}
\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
\end{equation*}
where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}. In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.

\subsection{Adjustment of the NVT Dynamics}

@@

Speak about the Generalized Equipartition Theorem....

@@

Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space. Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a point in phase space, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation}
\label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$). A generalized Liouville operator can then be defined as\cite{Sergi_2004}
\begin{equation*}
{\mathcal L} = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
\end{equation*}
